---
title: "Global Mental Health Disorders"
---

# What are the trends in Mental Health Disorders across the globe?

Mental Health refers to "mental well-being that enables people to cope with the stresses of life, realize their abilities, learn well and work well, and contribute to their community" (World Health Organization, 2022b). Mental health disorders are "characterized by a clinically significant disturbance in an individual’s cognition, emotional regulation, or behaviour" (World Health Organization, 2022a). There are a multitude of mental health disorders that exist. On this website, the data analysed are focused on 7 Mental Health Disorders — Alcohol use disorder, Anxiety disorder, Bipolar disorder, Depression, Drug use disorder, Eating disorder and Schizophrenia.

A study by researchers from Harvard Medical School and the University of Queensland have found that one out of every two people in the world will develop a mental health disorder in their lifetime (Queensland Brain Institute, 2023). This shows how prevalent mental health disorders have been in recent years. It is crucial to find out the main type and cause of mental health disorders in order to deal with them effectively. The current treatments used to deal with these disorder have not been very effective, hence data in this area of study will allow for greater ease of finding an effective treatment (Queensland Brain Institute, 2023).

This website provides a visualisation for Mental Health Disorders across the globe. It is meant to bring awareness to the mental health issues that people experience, albeit the small percentage it seems to take up of the whole population. It is important to note that not all who suffer from mental health disorders seek help and are being reported in the dataset.

# Dataset Used

The dataset used includes the columns: 

* Index: A unique number attached to each observation 
* Entity: Country name 
* Code: Country Code 
* Year: Year of data observed 
* Schizophrenia, Bipolar disorder, Eating disorders, Anxiety Disorders, Drug use Disorders, Depression, Alcohol Use Disorders: Percenatge of people reported to have the specific mental health disorder

```{r, eval=TRUE, include=FALSE}
library(tidyverse)
library(kableExtra)
library(geojsonio)
new_data <- read.csv("/Users/aliciaho/Documents/GitHub/aliiciaho.github.io/personal_app/refined_NMdata.csv")
countries <- geojsonio::geojson_read("https://raw.githubusercontent.com/lvictory/maps-data/master/world.geo.json", 
                                  what = "sp")
spaces_indices <- which(trimws(new_data$Code) == "")
cleaned_data <- new_data[-spaces_indices, ] %>% 
  filter(Year >= 1990)

# Cleaning table data
matching_values <- cleaned_data$Code %in% countries$id 
matched_dataset1 <- cleaned_data[matching_values, ]
```

```{r, eval=TRUE, echo=FALSE}
table <- head(new_data %>%
             rename(Index = index,
         Entity = Entity,
         Code = Code,
         Year = Year,
         Schizophrenia = Schizophrenia....,
         Bipolar_Disorder = Bipolar.disorder....,
         Eating_Disorder = Eating.disorders....,
         Anxiety_Disorder = Anxiety.disorders....,
         Druguse_Disorder = Drug.use.disorders....,
         Depression = Depression....,
         Alcoholuse_Disorder = Alcohol.use.disorders....))
table %>%
  knitr::kable() %>%
  kable_styling(full_width = FALSE, bootstrap_options = "hover")
```

There are a total of 168 countries that will be used in these visualisations. 

# Mental Health Disorder trend for each country
Below, you can find an interactive graph. Select the country you are interested in and its relevant graph will be produced. 

```{=html}
<iframe height="600" width="100%" frameborder="no" src="https://1ua3vn-alicia-ho.shinyapps.io/personal_app/"> </iframe>
```

# Alcohol Use Disorder
```{r, eval=TRUE, echo=FALSE}
alcohol <- matched_dataset1 %>% arrange(desc(Alcohol.use.disorders....)) %>% 
  filter(Year == 2017) %>%
  select(Entity, Code, Year, Alcohol.use.disorders....) %>%
  rename(
    Alcoholuse_Disorder = Alcohol.use.disorders.... 
  ) %>%
  slice(1:10)
Rank <- 1:10
alcohol <- alcohol[, !names(alcohol) %in% "Rank", drop = FALSE]

# Insert the new_column as the first column
alcohol <- cbind(Rank, alcohol)

alcohol %>%
  knitr::kable() %>%
  kable_styling(full_width = FALSE, bootstrap_options = "hover")
```
At a glance, it can be observed that most of the top 10 countries with a high percentage of their population suffering from alcohol use disorder originate from Eastern European countries. 

The country with the highest alcohol use disorders is Belarus with 5.34% of its population suffering from it in 2017. That is roughly 500,000 people out of the 9.459 million citizens of Belarus in 2017. Belarus has one of the highest alcohol consumption per capita in the world. 

```{r, eval=TRUE, echo=FALSE}
alcohol_low <- matched_dataset1 %>% arrange(Alcohol.use.disorders....) %>% 
  filter(Year == 2017) %>%
  select(Entity, Code, Year, Alcohol.use.disorders....) %>%
  rename(
    Alcoholuse_Disorder = Alcohol.use.disorders.... 
  ) %>%
  slice(1:10)
Rank <- c(168, 167, 165, 164, 163, 162, 161, 160, 159, 158)
alcohol_low <- alcohol_low[, !names(alcohol_low) %in% "Rank", drop = FALSE]
# Insert the new_column as the first column
alcohol_low <- cbind(Rank, alcohol_low)

alcohol_low %>%
  knitr::kable() %>%
  kable_styling(full_width = FALSE, bootstrap_options = "hover")
```

Most of these countries practice the religion of Islam, which considers drinking alcohol non-halal and it not allowed. This explains why these countries belong in the bottom 10 in terms of alcohol use disorders. 

# Limitations

The data that was curated only contains data up to 2017. Hence, it is not the most accurate representation of the current mental states of people in the world. With the emergence of the Covid-19 pandemic in 2019, research has shown that mental health has deteriorated immensely (cite). Ideally, the visualisations shown would include the most recent data for stakeholders to identify current poignant mental health issues that are on the rise.

Some countries are not included because of the mismatch of data between the countries in the global map and the countries in the mental health dataset.
