---
title: "Global Mental Health Disorders"
---

# What are the trends in Mental Health Disorders across the globe?

**Mental Health** refers to "mental well-being that enables people to cope with the stresses of life, realize their abilities, learn well and work well, and contribute to their community" (World Health Organization, 2022b). **Mental health disorders** are "characterized by a clinically significant disturbance in an individual’s cognition, emotional regulation, or behaviour" (World Health Organization, 2022a). There are a multitude of mental health disorders that exist. On this website, the data analysed are focused on 7 Mental Health Disorders — Alcohol use disorder, Anxiety disorder, Bipolar disorder, Depression, Drug use disorder, Eating disorder and Schizophrenia.

A study by researchers from Harvard Medical School and the University of Queensland have found that one out of every two people in the world will develop a mental health disorder in their lifetime (Queensland Brain Institute, 2023). This shows how prevalent mental health disorders have been in recent years. It is crucial to find out the main type and cause of mental health disorders in order to deal with them effectively. The current treatments used to deal with these disorder have not been very effective, hence data in this area of study will allow for greater ease of finding an effective treatment (Queensland Brain Institute, 2023).

This data story provides a visualisation for Mental Health Disorders across the globe from 1990-2017. It is meant to **bring awareness to the mental health issues that people experience**, albeit the small percentage it seems to take up of the whole population. Understanding the trends, disparities, and progress in mental health over time is crucial for **effective intervention and policy-making**. It is important to note that not all who suffer from mental health disorders seek help and are being reported in the dataset.

# Dataset Used

The dataset used includes the columns: 

* Index: A unique number attached to each observation 
* Entity: Country name 
* Code: Country Code 
* Year: Year of data observed 
* Schizophrenia, Bipolar disorder, Eating disorders, Anxiety Disorders, Drug use Disorders, Depression, Alcohol Use Disorders: Percenatge of people reported to have the specific mental health disorder

```{r, eval=TRUE, include=FALSE}
library(tidyverse)
library(kableExtra)
library(geojsonio)
library(ggplot2)
library(plotly)
new_data <- read.csv("/Users/aliciaho/Documents/GitHub/aliiciaho.github.io/personal_app/refined_NMdata.csv")
countries <- geojsonio::geojson_read("https://raw.githubusercontent.com/lvictory/maps-data/master/world.geo.json", 
                                  what = "sp")
spaces_indices <- which(trimws(new_data$Code) == "")
cleaned_data <- new_data[-spaces_indices, ] %>% 
  filter(Year >= 1990)

# Cleaning table data
matching_values <- cleaned_data$Code %in% countries$id 
matched_dataset1 <- cleaned_data[matching_values, ]
```

```{r, eval=TRUE, echo=FALSE}
table <- head(new_data %>%
             rename(Index = index,
         Entity = Entity,
         Code = Code,
         Year = Year,
         Schizophrenia = Schizophrenia....,
         Bipolar_Disorder = Bipolar.disorder....,
         Eating_Disorder = Eating.disorders....,
         Anxiety_Disorder = Anxiety.disorders....,
         Druguse_Disorder = Drug.use.disorders....,
         Depression = Depression....,
         Alcoholuse_Disorder = Alcohol.use.disorders....))
table %>%
  knitr::kable() %>%
  kable_styling(full_width = FALSE, bootstrap_options = "hover")
```

There are a total of **168 countries** that will be used in these visualisations.

# Overall trends of mental health disorder


### Which disorders should we be concerned about?
```{r, eval=TRUE, echo=FALSE}
median_values_2017 <- matched_dataset1 %>%
  filter(Year == 2017) %>%
  pivot_longer(cols = Schizophrenia....:Alcohol.use.disorders...., names_to = "disorder_type", values_to = "percentage") %>%
  group_by(disorder_type) %>%
  summarize(median_percentage = median(percentage)) %>% mutate(x_values = c(-0.32, -0.21, -0.11, 0, 0.11, 0.21, 0.32))

ggplot(matched_dataset1 %>%
         filter(Year ==2017) %>%
        pivot_longer(cols = Schizophrenia....:Alcohol.use.disorders...., names_to = "disorder_type", values_to = "percentage") %>%
        select(Year, disorder_type, percentage)) +
  aes(y = percentage, color = disorder_type) + 
  geom_boxplot() + 
  theme_bw() +
  labs(y = "Percentage of Population", title = "Spread of each global mental health disorder in 2017") +
  scale_color_discrete(
                name = "Type of mental health\ndisorder",
                labels = c("Alcohol use disorder", "Anxiety disorder", "Bipolar disorder", "Depression", "Drug use disorder", "Eating disorder", "Schizophrenia")
              ) +
  geom_label(data = median_values_2017, aes(label = paste(round(median_percentage, 3)), x = c(-0.32, -0.21, -0.11, 0, 0.11, 0.21, 0.32), y = median_percentage), 
            color = "black", size = 3, vjust = -0.3, show.legend = NA) +
  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(), axis.title.x = element_blank())
```

The graph above shows that the dominant mental health disorders that prevail in 2017 are **Anxiety disorders**, followed by **Depression** and **Alcohol Use Disorder**. 


```{r, eval=TRUE, echo=FALSE, message=FALSE}
invisible(overall_median <- matched_dataset1 %>%
  rename(index = index,
         Entity = Entity,
         Code = Code,
         Year = Year,
         Schizophrenia = Schizophrenia....,
         Bipolar_Disorder = Bipolar.disorder....,
         Eating_Disorder = Eating.disorders....,
         Anxiety_Disorder = Anxiety.disorders....,
         Druguse_Disorder = Drug.use.disorders....,
         Depression = Depression....,
         Alcoholuse_Disorder = Alcohol.use.disorders....) %>%
  pivot_longer(cols = Schizophrenia:Alcoholuse_Disorder, names_to = "disorder_type", values_to = "percentage") %>%
  group_by(Year, disorder_type) %>% summarise(median_year = median(percentage)) %>% ungroup()
)
ggplot(overall_median) +
  aes(x = Year, y = median_year, color = disorder_type) + 
  geom_point(shape="cross") +
  labs(y = "Median Value of % of Population across all countries", title = "Overall trend of median percentage values of mental health disorders worldwide \nfrom 1990-2017") +
  geom_line() +
  theme_bw() +
  theme(axis.text.x = element_text(colour = "grey20", size = 10, angle = 90, hjust = 0.5, vjust = 0.5) 
  ) +
   scale_color_discrete(
     name = "Type of mental health\ndisorder",
     labels = c("Alcohol use disorder", "Anxiety disorder", "Bipolar disorder", "Depression", "Drug use disorder", "Eating disorder", "Schizophrenia")
   ) +
    scale_y_continuous(minor_breaks = seq(1, 5, 0.2)) +
        scale_x_continuous(breaks = seq(1990, 2017, 1))

```
The trend of median percentages of mental health disorders across 1990-2017 shows that there are no significant increases or decreases. Although this shows that dealing with mental health disorders is not an issue of high priority, it also signifies that the **current methods of improving individuals' mental health are not effective**. 

# Which countries should pay greater attention to their citizens' mental health?

```{r, eval=TRUE, echo=FALSE}
x <- data.frame(
  Rank = c(1,2,3),
  Anxiety = c("New Zealand", "Norway", "Iran"),
  Depression = c("Greenland", "Lesotho", "Moroco"),
  Alcohol_Use_Disorder = c("Belarus", "Russia", "Estonia"),
  Drug_Use_Disorder = c("United States", "United Arab Emirates", "Libya"),
  Bipolar_Disorder = c("New Zealand", "Australia", "Brazil"),
  Eating_Disorder = c("Australia", "Luxembourg", "Spain"),
  Schizophrenia = c("Australia", "Netherlands", "New Zealand")
)
x %>%
  knitr::kable() %>%
  kable_styling(full_width = FALSE, bootstrap_options = "hover")
```
*Rank 1 indicates highest percentage of population with the mental health disorder*

Hover over the country of interest to find out the most recent mental health disorder data (2017). Toggle between the different mental health disorders to get a bird's eye view of data across the globe.

```{=html}
<iframe height="600" width="100%" frameborder="no" src="https://1ua3vn-alicia-ho.shinyapps.io/mapmap/"> </iframe>
```

# Mental Health Disorder trend for each country

Below, you can find an interactive graph. Select the country you are interested in and its relevant graph will be produced. 

```{=html}
<iframe height="600" width="100%" frameborder="no" src="https://1ua3vn-alicia-ho.shinyapps.io/personal_app/"> </iframe>
```

# Limitations

The data that was curated only contains data up to 2017. Hence, it is not the most accurate representation of the current mental states of people in the world. With the emergence of the Covid-19 pandemic in 2019, research has shown that mental health has deteriorated immensely (cite). Ideally, the visualisations shown would include the most recent data for stakeholders to identify current poignant mental health issues that are on the rise.

Some countries are not included because of the mismatch of data between the countries in the global map and the countries in the mental health dataset. (eg. Singapore)
